Transportation network with multiple lines

### 2022-04-11 multiple_lines/notes/latex/Notes_trajet.tex
Notes updated
Multi-lines notes updated

Note that if only $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ are given, a flow matrix $\mathbf{N}$ can be computed relatively to an \emph{orgin-destination affinity matrix} $\mathbf{S} = (s_{st})$, $0 \leq s_{st} \leq 1$, where $s_{st} = 1$ denote a perfect affinity and $s_{st} = 0$ no affinity, through

@@

From equation (\ref{sp_comp}), we see that 
\begin{equation}
\frac{\partial x^\text{sp}_{ij}}{\partial n_{st}} = \text{P}_{st}(i, j),
\end{equation}
which equal to $1$ if there is a unique shortest-path between $s$ and $t$ and $(i,j)$ belongs to this path.

@@

By setting the problem like that, we easily see that it is ill posed. Several solutions exists (e.g. units remain on the same line and follows a first-in/first-out scheme) and we need to add some hypotheses to restrain it.

\paragraph{Hypothesis 1:} Trajectories follow shortest-paths from origin $s$ to destination $t$. \\

\paragraph{Hypothesis 2:} The number of trajectories $n_{st}$ should be as close as possible to $\rho^\text{in}_s \rho^\text{out}_t s_{st}$, where $S = (s_{st})$ is a given affinity matrix between origin and destination nodes. \\

\subsection{Algorithm}
Set $\bm{\sigma}^{(1)}_\text{in} = \bm{\rho}_\text{in}$, $\bm{\sigma}^{(1)}_\text{out} = \bm{\rho}_\text{out}$, and $\mathbf{S}^{(1)} = S$. Until convergence, do:
\begin{enumerate}
	\item Compute $\mathbf{N}^{(i)} = \textbf{Diag}(\mathbf{a}^{(i)}) (\mathbf{S}^{(i)} + \epsilon)\textbf{Diag}(\mathbf{b}^{(i)})$ with iterative fitting, such that $\mathbf{N}^{(i)} \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(i)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(i)}_\text{out}  + \epsilon$.
	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(i)}$ with (\ref{sp_comp}).
	\item Compute $\bm{\sigma}^{(i+1)}_\text{in} = \bm{\rho}_\text{in} -  (\mathbf{X}^{(i)}_\text{B})^\top \mathbf{e}_n$ and $\bm{\sigma}^{(i+1)}_\text{out} = \bm{\rho}_\text{out} - \mathbf{X}^{(i)}_\text{B} \mathbf{e}_n$. 
	\item Set negative components of $\bm{\sigma}^{(i+1)}_\text{in}$ and $\bm{\sigma}^{(i+1)}_\text{out}$ to 0.
	\item Update $\mathbf{S}^{(i + 1)}$ according to excess flow.
\end{enumerate}

>>>>

Note that if only $\bm{\sigma}_\text{in}$ and $\bm{\sigma}_\text{out}$ are given, a flow matrix $\mathbf{N}$ can be computed relatively to an \emph{origin-destination affinity matrix} $\mathbf{S} = (s_{st})$, $0 \leq s_{st} \leq 1$, where $s_{st} = 1$ denote a perfect affinity and $s_{st} = 0$ no affinity, through

@@

From equation (\ref{sp_comp}), we see that 
\begin{equation}
\frac{\partial x^\text{sp}_{ij}}{\partial n_{st}} = \text{P}_{st}(i, j),
\end{equation}
which equal to $1$ if there is a unique shortest-path between $s$ and $t$ and $(i,j)$ belongs to this path. This equation means that if we multiply the flow by a factor $\alpha \geq 0$ on each $s, t$ which contains $(i, j)$ on their shortest-paths, the resulting flow on $(i, j)$ will also be multiplied by $\alpha$.

@@

By setting the problem like that, we easily see that it is ill posed. Several solutions exists (e.g. units remain on the same line and follow a first-in/first-out scheme) and we need to add some hypotheses to restrain it.

\paragraph{Hypothesis 1:} Trajectories in the network follow shortest-paths from origin $s$ to destination $t$. \\

\paragraph{Hypothesis 2:} The number of trajectories $\mathbf{N} = (n_{st})$ should be as close as possible to $s_{st}$, where $S = (s_{st})$ is a given affinity matrix between origin and destination nodes, in the sens that
\begin{equation}
	K(\mathbf{N} | \mathbf{S}) := \sum_{st} \frac{n_{st}}{n_{\bullet \bullet}} \log \left( \frac{n_{st}/n_{\bullet \bullet}}{s_{st} / s_{\bullet \bullet}} \right), \label{kl_div}
\end{equation}
i.e. the \emph{Kullback-Leibler divergence} between the probability of selecting an origin-destination path according to $\mathbf{N}$ relatively to the probabilty of selecting an origin-destination path according to $\mathbf{S}$, is minimum. Note that the divergence (\ref{kl_div}) is well defined only if $s_{st} > 0, \; \forall s, t$. \\

With these two additional hypotheses, we can find a solution with the following algorithm.

\subsection{Algorithm}
Set $\bm{\sigma}^{(1)}_\text{in} = \bm{\rho}_\text{in}$, $\bm{\sigma}^{(1)}_\text{out} = \bm{\rho}_\text{out}$, and $\mathbf{S}^{(1)} = \mathbf{S}$. Until convergence, do:
\begin{enumerate}
	\item Compute 
	\begin{equation}
		\mathbf{N}^{(\tau)} = \textbf{Diag}(\mathbf{a}^{(\tau)}) (\mathbf{S}^{(\tau)} + \epsilon)\textbf{Diag}(\mathbf{b}^{(\tau)}),
	\end{equation}
	with iterative fitting, such that $\mathbf{N}^{(\tau)} \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{in} + \epsilon$ and $\left(\mathbf{N}^{(\tau)}\right)^\top \mathbf{e}_n = \bm{\sigma}^{(\tau)}_\text{out}  + \epsilon$.

	\item Compute the associated shortest-path flow matrix $\mathbf{X}^{(\tau)}$ with (\ref{sp_comp}).

	\item Compute the vectors of \emph{between-lines flow entering and leaving each nodes}, i.e.
	\begin{align}
		\mathbf{x}^{(\tau)}_\text{B,in} &= (\mathbf{X}^{(\tau)}_\text{B})^\top \mathbf{e}_n \\
		\mathbf{x}^{(\tau)}_\text{B,out} &= \mathbf{X}^{(\tau)}_\text{B} \mathbf{e}_n
	\end{align}

	\item Compute the vectors of \emph{between-line allowed flow entering and leaving each nodes}, written resp. $\tilde{\mathbf{x}}^{(\tau)}_\text{B,in} = (\tilde{x}^{\text{B,in},(\tau)}_i)$ and $\tilde{\mathbf{x}}^{(\tau)}_\text{B,out} = (\tilde{x}^{\text{B,out},(\tau)}_i)$, with 
	\begin{align}
		\tilde{x}^{\text{B,in},(\tau)}_i &= \rho_i^\text{in} \phi \left( \frac{x^{\text{B,in},(\tau)}_i}{\rho_i^\text{in}} \right), \\
		\tilde{x}^{\text{B,in},(\tau)}_i &= \rho_i^\text{out} \phi \left( \frac{x^{\text{B,out},(\tau)}_i}{\rho_i^\text{out}} \right),
	\end{align}
	where $\phi(x)$ is a positive increasing function which should be the identity when $x \to 0$ and with $\phi(x) \leq 1$, $\forall x$. For example
	\begin{enumerate}
		\item $\phi(x) = \min(x, 1)$,
		\item $\phi(x) = \min(x, 1 - \exp(-\lambda x))$, with $\lambda > 0$ a parameter.
	\end{enumerate}

	\item Compute the \emph{between-lines allowed flow on edges}, noted $\widetilde{\mathbf{X}}_\text{B}^{(\tau)} = (\widetilde{x}_{ij}^{\text{B},(\tau)})$ with 
	\begin{equation}
		\widetilde{x}_{ij}^{\text{B},(\tau)} = \min \left( \frac{\tilde{x}^{\text{B,out},(\tau)}_i}{x^{\text{B,out},(\tau)}_i} , \frac{\tilde{x}^{\text{B,in},(\tau)}_j}{x^{\text{B,in},(\tau)}_j} \right)  x^{\text{B},(\tau)}_{ij}.
	\end{equation}
	\item Update the flow entering and leaving the network with
		\begin{align}
			\bm{\sigma}^{(\tau+1)}_\text{in} &= \bm{\rho}_\text{in} -  (\widetilde{\mathbf{X}}^{(\tau)}_\text{B})^\top \mathbf{e}_n, \\
			\bm{\sigma}^{(\tau+1)}_\text{out} &= \bm{\rho}_\text{out} -  \widetilde{\mathbf{X}}^{(\tau)}_\text{B} \mathbf{e}_n.
		\end{align}
	\item Compute the \emph{reducing factor matrix} $\mathbf{R}^{(\tau)} = (r^{(\tau)}_{st})$ with 
	\begin{equation}
		r^{(\tau)}_{st} = 1 - \max_{ij} \left( \text{P}_{st}(i, j) \frac{x_{ij}^{\text{B},(\tau)}  - \widetilde{x}_{ij}^{\text{B},(\tau)}}{x_{ij}^{\text{B},(\tau)} + \epsilon} \right).
	\end{equation}
	\item Update the origin-destination affinity matrix with
	\begin{equation}
		\mathbf{S}^{(\tau + 1)} = \mathbf{S}^{(\tau)} \odot \mathbf{R}^{(\tau)},
	\end{equation}
	where $\odot$ designate the Hadamard (component-wise) product of matrices. 
\end{enumerate}
